---
title: "Lab: Introduction to Regression"
output: html_notebook
---

# Load packages

```{r}
library(dplyr, warn.conflicts = FALSE)
library(readr)
library(broom)
library(ggplot2)
library(meantables)
library(freqtables)
```

# Overview

In this lab, we will explore some data collected from mothers and babies, which was indented to explore the occurrence of low birth weight. 

# Task 1

Please use the following links to download the low birth weight data to your computer and view the codebook:
 * [Click here to view/download the codebook](https://www.dropbox.com/s/n6za9yhtw44bh7f/Codebook lowbwt.pdf?dl=0). Look over the codebook to learn a little bit about the data you will be working with.
 * [Click here to view/download the data](https://www.dropbox.com/s/whel3kfsjswu2yy/lowbwt.rds?dl=0).
 
# Task 2

Import `lowbwt.rds` into your R global environment as `lowbwt`.

```{r}
lowbwt <- read_rds("/Users/bradcannell/Dropbox/Datasets/Low Birth Weight/lowbwt.rds")
```

# Task 3

Please view the structure of `lowbwt` using the `str()` function or the `glimpse()` function.

```{r}
glimpse(lowbwt)
```

# Question 1

* The `lowbwt` data frame contains ____ rows and ____ columns.

* The `lowbwt` data frame contains 189 rows and 15 columns.

# Task 4

Please fit a linear regression model to the `lowbwt` data frame using `weight of mother at last menstrual period in pounds` as the regressand and the `age of the mother in years` as the regressor.

```{r}
glm(
  lwt ~ age,
  family = gaussian(link = "identity"),
  data   = lowbwt
)
```

# Question 2

What is the mean value of the `weight of mother at last menstrual period in pounds` when the `mother’s age` is 0? (round to the nearest integer – no decimal places)

The mean value of the weight of mother at last menstrual period in pounds when the mother’s age is 0 is 106.

# Task 5

Please fit a linear regression model to the `lowbwt` data frame using `weight of mother at last menstrual period in pound`s as the regressand and `mother’s smoking status` during pregnancy as the regressor.

```{r}
glm(
  lwt ~ smoke_f,
  family = gaussian(link = "identity"),
  data   = lowbwt
)
```

# Question 3

* What is the mean value of the `weight of mother at last menstrual period in pounds` when the `mother smoked during pregnancy`? (round to the nearest integer – no decimal places)

```{r}
130.896 - 2.761
```

* The mean value of the `weight of mother at last menstrual period in pounds` when the `mother smoked during pregnancy` was 128. (rounded to the nearest integer – no decimal places)

# Task 6

Please fit a logistic regression model to the `lowbwt` data frame using `low birth weight` as the regressand and `age of mother in years` as the regressor.

```{r}
glm(
  low  ~ age,
  family = binomial(link = "logit"),
  data   = lowbwt
) %>% 
  tidy(exp = TRUE)
```

# Question 4

* On average, babies had _____ times the odds of `low birth weight` for each one-year increase in mother's age.(round to the hundredths place – two decimals)

* On average, babies had 0.95 times the odds of `low birth weight` for each one-year increase in mother's age.

# Task 7

Please fit a logistic regression model to the `lowbwt` data frame using `low birth weight` as the regressand and `mother’s smoking status during pregnancy` as the regressor.

```{r}
glm(
  low  ~ smoke_f,
  family = binomial(link = "logit"),
  data   = lowbwt
) %>% 
  tidy(exp = TRUE)
```

# Question 5

* On average, `low birth weight` babies had _____ times the odds of having a `mother who smoked` compared to babies that were not born with a `low birth weight`. (round to the hundredths place – two decimals)

* On average, `low birth weight` babies had 2.02 times the odds of having a `mother who smoked` compared to babies that were not born with a `low birth weight`. (rounded to the hundredths place – two decimals)

# Task 8

Please fit a Poisson regression model to the `lowbwt` data frame using `history of premature labor` as the regressand and `age of mother in years` as the regressor.

```{r}
glm(
  ptl  ~ age,
  family = poisson(link = "log"),
  data   = lowbwt
) %>% 
  tidy(exp = TRUE)
```

# Question 6

* Mother's reported _____ times the number of `premature deliveries`, on average, for each one-year increase in `age`. (rounded to the hundredths place – two decimals)

* Mother's reported 1.03 times the number of `premature deliveries` for each one-year increase in `age`. (rounded to the hundredths place – two decimals)

# Task 9

Please fit a Poisson regression model to the `lowbwt` data frame using `history of premature labor` as the regressand and `mother’s smoking status during pregnancy` as the regressor.

```{r}
glm(
  ptl  ~ smoke_f,
  family = poisson(link = "log"),
  data   = lowbwt
) %>% 
  tidy(exp = TRUE)
```

# Question 7

* Mother's who `smoked during pregnancy` reported an average of _____ times the number of `premature deliveries` compared to mothers who did not smoke during pregnancy. (round to the hundredths place – two decimals)

* Mother's who `smoked during pregnancy` reported an average of 2.55 times the number of `premature deliveries` compared to mothers who did not smoke during pregnancy. (rounded to the hundredths place – two decimals)