---
title: "Lab Warm-up: Measures of Association"
output: html_notebook
---

# Load packages

```{r}
library(dplyr, warn.conflicts = FALSE)
library(freqtables)
```

# Create 2x2 contingency tables

There are many difference possible ways to create 2x2 contingency tables in R. I prefer building them as a matrix object. We haven't talked much about matrix objects up the this point, and for the most part, we don't need to use them very often.

## Matrix dimensions

In R, a matrix is a vector with 2 dimensions. What does that mean?

Here is a character vector.

```{r}
v <- c("a", "b", "c", "d")
v
```

If we want to select element "a", we can use bracket notation.

```{r}
v[1]
```

Here is a matrix with character values. Remember, all elements of a vector must be of the same time (character, numeric, etc.), and a matrix *is* a vector. So, all elements of a matrix must be of the same type.

```{r}
mat <- matrix(v)
mat <- matrix(c("a", "b", "c", "d"))
mat
```

Notice that by default, this matrix has 4 rows and 1 column.

And if we want to select the "a" element, we add both coordinates (dimensions) to our bracket notation. We do so by typing the matrix name followed by square brakets that contain the row we want, a comma, and the column we want. So, `matrix_name[row, column]`.

```{r}
mat[1, 1]

# Note. In this case, mat[1] would return the same result, but that won't be 
# true when we have more complex matrices.
```

## Matrix to 2x2 contingency tables

Now, to make our matrix look and behave like a 2x2 table, we need our matrix to have 2 rows and 2 columns instead of 4 rows and 1 column. First, we can adjust either the `ncol` or `nrow` argument to the `matrix()` function.

```{r}
matrix_ct <- matrix(
  c("a", "b", "c", "d"), # Enter values in this order
  ncol = 2
)

matrix_ct
```

```{r}
matrix_ct <- matrix(
  c("a", "b", "c", "d"), # Enter values in this order
  nrow = 2
)

matrix_ct
```

But, that still doesn't quite give us the result we want. By default, the `byrow` argument to the `matrix()` function is set to `FALSE`. This means that R will fill in the matrix values starting at location [1, 1] and filling down columns. We want R to fill in the matrix values starting at location [1, 1] and filling across rows. We do that by changing the default `byrow = FALSE` to `byrow = TRUE`.

```{r}
matrix_ct <- matrix(
  c("a", "b", "c", "d"), # Enter values in this order
  nrow = 2,
  byrow = TRUE
)

matrix_ct
```

Now, we have a matrix object that looks and behaves like a 2x2 contingency table.

## Add row and column names

To make our matrix even easier to read, let's go ahead and add row and column names.

```{r}
dimnames(matrix_ct) <- list(
  Exposure = c("Yes", "No"), # Row names
  Outcome  = c("Yes", "No")  # Then column names
)

matrix_ct
```

## Add margins

Finally, let's add our marginal totals.

Because we can't add letters together, let's use the numbers from our slide example.

```{r}
matrix_ct <- matrix(
  c(a = 3, b = 2, c = 3, d = 2),
  ncol = 2,
  byrow = TRUE
)

dimnames(matrix_ct) <- list(
  Exposure = c("Yes", "No"), # Row names
  Outcome  = c("Yes", "No")  # Then column names
)

matrix_ct
```

And now let's add the marginal totals. We could do this manually.

```{r}
matrix_ct[, 1]
```

```{r}
matrix_ct[, 2]
```

```{r}
row_total <- matrix_ct[, 1] + matrix_ct[, 2]
row_total
```

Let's band those values as a new column in our matrix with the `cbind()` function.

```{r}
matrix_ct_row_total <- cbind(matrix_ct, row_total)
matrix_ct_row_total
```

And then repeat for the column totals.

```{r}
matrix_ct_row_total[1, ]
```

```{r}
matrix_ct_row_total[2, ]
```

```{r}
col_total <- matrix_ct_row_total[1, ] + matrix_ct_row_total[2, ]
col_total
```

Notice we use `rbind()` instead of `cbind()` this time.

```{r}
matrix_ct_margins <- rbind(matrix_ct_row_total, col_total)
matrix_ct_margins
```

We could also just use the `addmargins()` function.

```{r}
matrix_ct_margins <- addmargins(matrix_ct)
matrix_ct_margins
```

# Incidence proportion ratio

-   Relative risk
-   Risk ratio

$$\frac{Incidence\, proportion\, in\, the\, exposed}{Incidence\, proportion\, in\, the\, unexposed}$$

$$\frac{\frac{a}{(a+b)}}{\frac{c}{(c+d)}}$$

```{r}
matrix_ct <- matrix(
  c(a = 3, b = 2, c = 2, d = 2),
  ncol = 2,
  byrow = TRUE
)

matrix_ct <- addmargins(matrix_ct)

dimnames(matrix_ct) <- list(
  Exposure = c("Yes", "No", "col_sum"), # Row names
  Disease  = c("Yes", "No", "row_sum")  # Then column names
)

matrix_ct
```

```{r}
incidence_prop <- matrix_ct[, "Yes"] / matrix_ct[, "row_sum"]
incidence_prop
```

```{r}
matrix_ct <- cbind(matrix_ct, incidence_prop)
matrix_ct
```

Calculate our incidence proportion ratio

```{r}
ipr <- matrix_ct["Yes", "incidence_prop"] / matrix_ct["No", "incidence_prop"]
ipr
```

## Interpretation

1.  Among the members of our population, those who were exposed had 1.2 times the incidence of disease compared to those who were not exposed over the 12 months of follow-up period.

2.  Among the members of our population, those who were exposed had 1.2 times the risk of disease compared to those who were not exposed over the 12 months of follow-up period.

# Incidence proportion difference

-   Risk difference

$${Incidence\, proportion\, in\, the\, exposed}-{Incidence\, proportion\, in\, the\, unexposed}$$

$$\frac{a}{(a+b)}-\frac{c}{(c+d)}$$

```{r}
matrix_ct
```

Calculate our incidence risk difference

```{r}
ird <- matrix_ct["Yes", "incidence_prop"] - matrix_ct["No", "incidence_prop"]
ird
```

## Interpretation

1.  Among the members of our population, those who were exposed had 0.1 additional cases of disease per person compared to those who were not exposed over the 12 months of follow-up period.

```{r}
ird * 100
```

2.  Among the members of our population, those who were exposed had 10 additional cases of disease per 100 people compared to those who were not exposed over the 12 months of follow-up period.

**Tip 1:** If you have the incidence proportion expressed as percentage, convert it to convenient fractions so that you can express it as the additional risk in a group of people who have the exposure.

**Tip 2:** Focus your interpretation on the additional risk in the exposed group.

Example: "There were 10 additional cases of disease per 100 people in the group that was exposed, compared to the group without exposure."

**Tip 3:** Don't forget to specify the time interval when using incidence proportion.

Example: "In the group that failed to adhere closely to the Mediterranean diet there were 120 excess deaths per 1,000 men during the two year period of observation compared to the group that did adhere to the Mediterranean diet."

# Incidence odds ratio

-   Odds ratio

$$\frac{Incidence\, odds\, in\, the\, exposed}{Incidence\, odds\, in\, the\, unexposed}$$

$$\frac{\frac{a}{b}}{\frac{c}{d}}$$

```{r}
matrix_ct
```

Calculate odds

```{r}
incidence_odds <- matrix_ct[, "incidence_prop"] / (1 - matrix_ct[, "incidence_prop"])
incidence_odds
```

```{r}
incidence_odds <- matrix_ct[, "Yes"] / matrix_ct[, "No"]
incidence_odds
```

```{r}
matrix_ct <- cbind(matrix_ct, incidence_odds)
matrix_ct
```

Calculate our incidence odds ratio

```{r}
ior <- matrix_ct["Yes", "incidence_odds"] / matrix_ct["No", "incidence_odds"]
ior
```

## Interpretation

1. Among the members of our population, those who were exposed had 1.5 times the odds of incident disease compared to those who were not exposed over the 12 months of follow-up period.
