---
title: "Lab: Effect Measure Modification"
format:
  html:
    embed-resources: true
---

# Load packages

Remember, it's considered a best practice to load all the packages that your file will use up at the top of the file. If you have not yet installed a package, you will need to do so by running `install.packages("package name")` in the R console. For example, to install the `dplyr` package, you would run `install.packages("dplyr")`. However, you typically **do not** want to type the code to install the package here in your Quarto file because you only need to install the package once on a given computer. Not every time you run your R code.

```{r}
#| label: load-packages
library(dplyr, warn.conflicts = FALSE) # The "warn.conflicts" part is optional
```


# Overview

In this lab, we will work through several scenarios that are intended to help us better understand effect-measure modification.


## Question 1

In a prospective study of the relationship between hepatitis C virus (HCV) and newly developed hepatocellular carcinoma, the investigators were interested in a potential modifying effect of alcohol consumption. The following table is based on this study's results:

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY |
|------------------|----------|----------------|----------------------------|
| Absent           | Negative | 8968           | 78.7                       |
|                  | Positive | 2352           | 127.1                      |
| Present          | Negative | 461            | 309.7                      |
|                  | Positive | 90             | 384.9                      |


Using **Absent–Negative** as the reference category, please calculate the incidence rate ratio for participants who did not drink alcohol and had HCV, participants who drank alcohol and did not have HCV, and participants who drank alcohol and had HCV. Please round your answers to two decimal places, and use them to fill in the table below.

```{r}
# Calculate incidence rate ratios
round(127.1 / 78.7, 2)
round(309.7 / 78.7, 2)
round(384.9 / 78.7, 2)
```

* Absent-Positive = 1.61
* Present-Negative = 3.94
* Present-Positive = 4.89

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Ratio (IRR) |
|------------------|----------|----------------|----------------------------|----------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                  |
|                  | Positive | 2352           | 127.1                      | 1.61                       |
| Present          | Negative | 461            | 309.7                      | 3.94                       |
|                  | Positive | 90             | 384.9                      | 4.89                       |

## Notes for students

1. The layout of the table above is different than the layout of the typical 2x2 contingency tables we usually work with. Additionally, we are concerned with two exposures (i.e., alcohol drinking and HCV) in this scenario instead of just one exposure. Even so, we can calculate the incidence rate ratio in exactly the same way we always have -- the incidence rate in the exposed divided by the incidence rate in the unexposed (or reference category). We will just need to do the calculation three times because we have three levels of exposure (i.e., Absent-Positive, Present-Negative, and Present-Positive). 

2. In this case, we are told that **Absent–Negative** is the reference category. Therefore, the ratio of the incidence rate in the exposed (to HCV, but not alcohol) to the unexposed is 127.1 / 78.7 = 1.61. In other words, among the participants in this study, those who were exposed to HCV, but not alcohol drinking, had 1.61 times the rate of hepatocellular carcinoma compared to those who were not exposed to either HCV or alcohol drinking during the follow-up period of the study.

3. The same logic then applies to each of the other exposure groups.

## Question 2

In a prospective study of the relationship between hepatitis C virus (HCV) and newly developed hepatocellular carcinoma, the investigators were interested in a potential modifying effect of alcohol consumption. Using the table of incidence rate ratios you created in a previous question, please calculate the expected joint incidence rate ratio (multiplicative model) in the exposed. Please round your answer to two decimal places.

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Ratio (IRR) |
|------------------|----------|----------------|----------------------------|----------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                  |
|                  | Positive | 2352           | 127.1                      | 1.61                       |
| Present          | Negative | 461            | 309.7                      | 3.94                       |
|                  | Positive | 90             | 384.9                      | 4.89                       |

```{r}
# Calculate the expected joint incidence rate ratio (multiplicative model) 
# in the exposed.
round(1.61 * 3.94, 2)
```

* 6.34

## Notes for students

1. When there is no effect-measure modification, we expect the combined effects of our two exposures to simply equal the combination of their individual effects. On a multiplicative scale, we combine the individual effects by multiplication.

2. The individual effect of HCV (i.e., in the absence of alcohol drinking) is in row two of the table above -- 1.61.

3. The individual effect of alcohol drinking (i.e., in the absence of HCV) is in row three of the table above -- 3.94. 

4. Therefore, the expected combined effect of alcohol drinking and HCV is 1.61 * 3.94 = 6.34.

## Question 3

Based on your expert knowledge of alcohol consumption, HCV, and hepatocellular carcinoma, a difference of 1 unit or more on the incidence rate ratio scale would constitute a meaningful difference in risk. Assuming there is no random variability, do the results of this study provide evidence that alcohol consumption meaningfully modifies the relationship between hepatitis C virus exposure and newly developed hepatocellular carcinoma on the multiplicative scale? The table of incidence rate ratios you created in a previous question is provided below.

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Ratio (IRR) |
|------------------|----------|----------------|----------------------------|----------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                  |
|                  | Positive | 2352           | 127.1                      | 1.61                       |
| Present          | Negative | 461            | 309.7                      | 3.94                       |
|                  | Positive | 90             | 384.9                      | 4.89                       |

```{r}
# Calculate the difference between the expected joint IRR and observed joint IRR
6.34 - 4.89
```

* Yes. There is evidence of meaningful effect-measure modification on the multiplicative scale because the expected joint incidence rate ratio (6.34) is different from the observed joint incidence ratio (4.89) by 1.45 units. Based on your expert knowledge of alcohol consumption, HCV, and hepatocellular carcinoma, you know that a difference of 1 unit or more on the incidence rate ratio scale constitutes a meaningful difference in risk.

## Question 4

See Canvas. There are no R calculations for this question.

## Question 5

In a prospective study of the relationship between hepatitis C virus (HCV) and newly developed hepatocellular carcinoma, the investigators were interested in a potential modifying effect of alcohol consumption. The following table is based on this study's results:

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY |
|------------------|----------|----------------|----------------------------|
| Absent           | Negative | 8968           | 78.7                       |
|                  | Positive | 2352           | 127.1                      |
| Present          | Negative | 461            | 309.7                      |
|                  | Positive | 90             | 384.9                      |


Using **Absent–Negative** as the reference category, please calculate the incidence rate _difference_ for participants who did not drink alcohol and had HCV, participants who drank alcohol and did not have HCV, and participants who drank alcohol and had HCV. Please round your answers to one decimal place, and use them to fill in the table below.

```{r}
# Calculate incidence rate differences
127.1 - 78.7
309.7 - 78.7
384.9 - 78.7
```

* Absent-Positive = 48.4
* Present-Negative = 231.0
* Present-Positive = 306.2

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Difference (IRD) |
|------------------|----------|----------------|----------------------------|---------------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                       |
|                  | Positive | 2352           | 127.1                      | 48.4                            |
| Present          | Negative | 461            | 309.7                      | 231.0                           |
|                  | Positive | 90             | 384.9                      | 306.2                           |

## Notes for students

1. The layout of the table above is different than the layout of the typical 2x2 contingency tables we usually work with. Additionally, we are concerned with two exposures (i.e., alcohol drinking and HCV) in this scenario instead of just one exposure. Even so, we can calculate the incidence rate difference in exactly the same way we always have -- the incidence rate in the exposed minus the incidence rate in the unexposed (or reference category). We will just need to do the calculation three times because we have three levels of exposure (i.e., Absent-Positive, Present-Negative, and Present-Positive). 

2. In this case, we are told that **Absent–Negative** is the reference category. Therefore, the difference between the incidence rate in the exposed (to HCV, but not alcohol) and the unexposed is 127.1 - 78.7 = 48.4. In other words, among the participants in this study, those who were exposed to HCV, but not alcohol drinking, experienced 48.4 additional cases of hepatocellular carcinoma per 100,000 person-years compared to those who were not exposed to either HCV or alcohol drinking during the follow-up period of the study.

3. The same logic then applies to each of the other exposure groups.

## Question 6

In a prospective study of the relationship between hepatitis C virus (HCV) and newly developed hepatocellular carcinoma, the investigators were interested in a potential modifying effect of alcohol consumption. Using the table of incidence rate differences you created in a previous question, please calculate the expected joint incidence rate difference (additive model) in the exposed. Please round your answer to one decimal place.

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Difference (IRD) |
|------------------|----------|----------------|----------------------------|---------------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                       |
|                  | Positive | 2352           | 127.1                      | 48.4                            |
| Present          | Negative | 461            | 309.7                      | 231.0                           |
|                  | Positive | 90             | 384.9                      | 306.2                           |

```{r}
# Calculate the expected joint incidence rate difference (additive model) 
# in the exposed.
round(48.4 + 231.0, 1)
```

* 279.4

## Notes for students

1. When there is no effect-measure modification, we expect the combined effects of our two exposures to simply equal the combination of their individual effects. On an additive scale, we combine the individual effects by addition.

2. The individual effect of HCV (i.e., in the absence of alcohol drinking) is in row two of the table above -- 48.4.

3. The individual effect of alcohol drinking (i.e., in the absence of HCV) is in row three of the table above -- 231.0. 

4. Therefore, the expected combined effect of alcohol drinking and HCV is 48.4 + 231.0 = 279.4.

## Question 7

Based on your expert knowledge of alcohol consumption, HCV, and hepatocellular carcinoma, a difference of 20 cases or more per 100,000 person-years would constitute a meaningful difference in risk. Assuming there is no random variability, do the results of this study provide evidence that alcohol consumption meaningfully modifies the relationship between hepatitis C virus exposure and newly developed hepatocellular carcinoma on the additive scale? The table of incidence rate ratios you created in a previous question is provided below.

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Difference (IRD) |
|------------------|----------|----------------|----------------------------|---------------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                       |
|                  | Positive | 2352           | 127.1                      | 48.4                            |
| Present          | Negative | 461            | 309.7                      | 231.0                           |
|                  | Positive | 90             | 384.9                      | 306.2                           |

```{r}
# Calculate the difference between the expected joint IRD and observed joint IRD
306.2 - 279.4
```

* Yes. There is evidence of meaningful effect-measure modification on the additive scale because the expected joint incidence rate difference (279.4) is different from the observed joint incidence rate difference (306.2) by 26.8 cases per 100,000 person-years. Based on your expert knowledge of alcohol consumption, HCV, and hepatocellular carcinoma, you know that a difference of 20 cases or more per 100,000 person-years constitutes a meaningful difference in risk.

## Question 8

See Canvas. There are no R calculations for this question.

## Question 9

In a prospective study of the relationship between hepatitis C virus (HCV) and newly developed hepatocellular carcinoma, the investigators were interested in a potential modifying effect of alcohol consumption. The following table is based on this study's results:

| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY |
|------------------|----------|----------------|----------------------------|
| Absent           | Negative | 8968           | 78.7                       |
|                  | Positive | 2352           | 127.1                      |
| Present          | Negative | 461            | 309.7                      |
|                  | Positive | 90             | 384.9                      |

You previously determined that there was negative effect-measure modification on the multiplicative scale and positive effect-measure modification on the additive scale by making a comparison between observed and expected joint effects. Now, let's confirm your conclusions by exploring the homogeneity of effects across levels of the potential effect modifier (alcohol drinking). Please start by filling in the table below.

```{r}
# Calculate the incidence rate ratio among participants who did not drink alcohol
round(127.1 / 78.7, 2)
```

```{r}
# Calculate the incidence rate difference among participants who did not drink alcohol
round(127.1 - 78.7, 2)
```

```{r}
# Calculate the incidence rate ratio among participants who drank alcohol
round(384.9 / 309.7, 2)
```

```{r}
# Calculate the incidence rate difference among participants who drank alcohol
round(384.9 - 309.7, 2)
```


| Alcohol drinking | HCV      | No. of persons | Incidence rates/100,000 PY | Incidence Rate Ratio (IRR) | Incidence Rate Difference (IRD) |
|------------------|----------|----------------|----------------------------|----------------------------|---------------------------------|
| Absent           | Negative | 8968           | 78.7                       | Reference                  | Reference                       |
|                  | Positive | 2352           | 127.1                      | 1.61                       | 48.40                           |
| Present          | Negative | 461            | 309.7                      | Reference                  | Reference                       |
|                  | Positive | 90             | 384.9                      | 1.24                       | 75.20                           |

## Notes for students

1. The layout of the table above is different than the layout of the typical 2x2 contingency tables we usually work with. Additionally, we are concerned with two exposures (i.e., alcohol drinking and HCV) in this scenario instead of just one exposure. Even so, we can calculate the incidence rate ratio and incidence rate difference in exactly the same way we always have -- the incidence rate in the exposed divided by the incidence rate in the unexposed (or reference category). We will just need to do the calculation twice -- once for people who did not drink alcohol and once for people who did drink alcohol.

2. Within each level of alcohol drinking, HCV Negative is the reference group. Therefore, the ratio of the incidence rate in the exposed (to HCV) to the unexposed is 127.1 / 78.7 = 1.61 among participants who did not drink alcohol, and 384.9 / 309.7 = 1.24 among participants who did drink alcohol.

3. Likewise, the difference between the incidence rate in the exposed (to HCV) and the unexposed is 127.1 - 78.7 = 48.40 cases per 100,000 person-years among participants who did not drink alcohol, and 384.9 - 309.7 = 75.20 cases per 100,000 person-years among participants who did drink alcohol.

## Question 10

See Canvas. There are no R calculations for this question.
