---
title: "Confounding - Questions in Development"
format:
  html:
    embed-resources: true
---


# Load packages

```{r}
#| label: load-packages
library(dplyr, warn.conflicts = FALSE) # The "warn.conflicts" part is optional
library(ggdag, warn.conflicts = FALSE)
library(ggplot2)
```

## Change in estimate criteria

<!-- See slide 18 in the lab warm-up -->

A research study seeks to estimate the causal effect of yellow fingers on lung cancer. The true causal DAG for the relationship between yellow fingers and lung cancer is shown below.

<!-- Change the variables -->

```{r}
yellow_fingers_dag <- dagify(
  y ~ z, # The form is effect ~ cause
  x ~ z,
  # Optionally add labels to make the DAG easier to read
  labels = c(
    "x" = "Yellow fingers",
    "z" = "Smoking",
    "y" = "Lung cancer"
  ),
  # Optionally add coordinates to control the placement of the nodes on the DAG
  coords = list( 
    x = c(x = 1, z = 2, y = 3),
    y = c(x = 0, z = 1, y = 0)
  )
)

# Set seed to reproduce label locations
set.seed(2)

# Plot the dag
plot_yellow_fingers_dag <- ggdag(yellow_fingers_dag, text = FALSE, use_labels = "label") +
  theme_dag()

plot_yellow_fingers_dag
```

```{r}
#| eval: false
ggsave("plot_yellow_fingers_dag.png", width = 13.3, height = 7.5)
```

## Traditional criteria

A research study seeks to estimate the causal effect of education on breast cancer. The true causal DAG for the relationship between education and breast cancer is shown below.

```{r}
edu_bc_dag <- m_bias(
  x = "Education", 
  y = "Breast cancer", 
  a = "Income during childhood",
  b = "Genetic risk \nfor breast cancer", 
  m = "Mother's breast cancer"
)

# Set seed to reproduce label locations
set.seed(1)

plot_edu_bc_dag <- ggdag(edu_bc_dag, text = FALSE, use_labels = "label") +
  theme_dag()

plot_edu_bc_dag
```

```{r}
#| eval: false
ggsave("plot_edu_bc_dag.png", width = 13.3, height = 7.5)
```
