---
title: "table3-3"
output: html_notebook
---

```{r}
library(epitools)
```

```{r}
RRtable<-matrix(c(9970,9820,30,180),nrow = 2, ncol = 2)
```

```{r}
riskratio.wald(RRtable)
```

# Brad's notes

I think it would make more sense to flip the numbers in "Disease1" and "Disease2" (I really dislike the naming conventions this package is using!). 

It might also be helpful to name the rows and columns of our 2x2 table (matrix).

For example:

```{r}
table_3_3 <- matrix(
  c(180, 9820, 30, 9970),
  nrow = 2,
  byrow = TRUE
)

table_3_3
```

```{r}
colnames(table_3_3) <- c("MI+", "MI-")
rownames(table_3_3) <- c("severe hypertension", "normal")

table_3_3
```

Now, we can pass the matrix to `riskratio.wald(RRtable)`

```{r}
riskratio.wald(table_3_3)
```

Why is it using the exposed group as the reference?

Let's try writing our own function:

First, convert our matrix to a table object and add marginal totals.

```{r}
table_3_3 <- table_3_3 %>%
  as.table() %>% 
  addmargins()

table_3_3
```

Add the row and column names again.

```{r}
colnames(table_3_3) <- c("MI+", "MI-", "total")
rownames(table_3_3) <- c("severe hypertension", "normal", "total")

table_3_3
```

Figure out the calculations we want to use.

```{r}
# First number is the row, second number is the column
incidence_expoosed <- table_3_3[1, 1] / table_3_3[1, 3]
incidence_unexposed <- table_3_3[2, 1] / table_3_3[2, 3]
rr <- incidence_expoosed / incidence_unexposed
rr
```

Then add to function.

```{r}
risk_ratio <- function(.table) {
  incidence_expoosed  <- .table[1, 1] / .table[1, 3]
  incidence_unexposed <- .table[2, 1] / .table[2, 3]
  rr <- incidence_expoosed / incidence_unexposed
  rr
}
```

Test it out.

```{r}
risk_ratio(table_3_3)
```

Possible further improvements:

* Make a wrapper function that simplifies the creation of the 2x2 table, or possibly move that functionality inside of `risk_ratio()`?
* Make new function to calculate confidence intervals and conduct hypothesis test, or or possibly move that functionality inside of `risk_ratio()`?
* Make it more general so that it can accept tables of any size -- not just 2x2.
* Others?





