---
title: "Chocolate Study"
output: html_notebook
---

# Overview

The NTRHD is part of a study about the association between chocolate eating habits and academic ability. The study took place at multiple schools, chosen at random, from around the region. Graduate students administered the surveys to students at each school, recorded survey responses electronically, and then emailed them to us. Now, we need your help importing the data into R, doing a little bit of data cleaning, and saving the data again once it’s ready for analysis. 

```{r message=FALSE}
library(dplyr, warn.conflicts = FALSE)
library(readxl)
library(readr)
library(haven)
```

# Task 1 

Click the links below to download the raw data files to your computer: 

* [chocolate study 1.xls](https://www.dropbox.com/s/9riqfhtf582o2az/Chocolate%20Study%201.xls?dl=1)   
* [chocolate study 2.csv](https://www.dropbox.com/s/ggln051redw1g98/Chocolate%20Study%202.csv?dl=1)
* [chocolate study 3.txt](https://www.dropbox.com/s/90ndmdc463ui88j/Chocolate%20Study%203.txt?dl=1)

# Task 2

Import `Chocolate Study 1.xls` into your R global environment as `choco_1`.

## Question:

1. Which of the following code chunks will correctly import the `Chocolate Study 1.xls` data into your R global environment as `choco_1`?

```{r eval=FALSE}
# Incorrect: chocolate_study_1.xls is not a csv file.
choco_1 <- read_csv("Chocolate Study 1.xls")
```

```{r eval=FALSE}
# Incorrect: readxl is the name of an R package, not a function.
choco_1 <- readxl("Chocolate Study 1.xls")
```

```{r eval=FALSE}
# Incorrect: readr is the name of an R package, not a function.
choco_1 <- readr("Chocolate Study 1.xls")
```

```{r eval=FALSE}
# Correct.
choco_1 <- read_excel("Chocolate Study 1.xls")
```

```{r}
# For Dr. Cannell's computer.
choco_1 <- read_excel("/Users/bradcannell/Dropbox/Datasets/Chocolate/Chocolate Study 1.xls")
```

# Task 3 

View the structure of `choco_1` using the `str()` function.

```{r}
str(choco_1)
```

## Question:

2. The `choco_1` data frame contains ____ rows and ____ columns.

* The `choco_1` data frame contains 50 rows and 21 columns.

# Task 4

Import `Chocolate Study 2.csv` into your R global environment as `choco_2`.

## Question:

3. Which of the following code chunks will correctly import the `Chocolate Study 2.csv` data into your R global environment?

```{r eval=FALSE}
# Incorrect: Although you can open chocolate_study_1.csv in Excel, it doesn't have a .xls or .xlsx file extension, which is what read_excel expects.
choco_2 <- read_excel("Chocolate Study 2.csv")
```

```{r eval=FALSE}
# Incorrect: There is a read.csv() function that will import this data. However, it isn't in the readr package. The double colon syntax used in the code readr::read.csv() tells R to look for the read.csv() function in the readr package, which doesn't exist. The read_csv() function is a base R function. 
choco_2 <- readr::read.csv("Chocolate Study 2.csv")
```

```{r eval=FALSE}
# Correct. However, I prefer read_csv() over read.csv().
choco_2 <- read.csv("Chocolate Study 2.csv")
```

```{r eval=FALSE}
# Correct.
choco_2 <- read_csv("Chocolate Study 2.csv")
```

```{r}
# For Dr. Cannell's computer.
choco_2 <- read_csv("/Users/bradcannell/Dropbox/Datasets/Chocolate/Chocolate Study 2.csv")
```

# Task 5 

View the structure of `choco_2` using the `str()` function.

```{r}
str(choco_2)
```

## Question:

4. The `choco_2` data frame contains ____ rows and ____ columns.

* The `choco_2` data frame contains 20 rows and 21 columns.

# Task 6 

Import `Chocolate Study 3.txt` into your R global environment as `choco_3`. Make sure to tell R how to correctly identify missing values in the raw data.

## Question:

5. Which of the following code chunks will correctly import the `Chocolate Study 3.txt` data into your R global environment?

```{r eval=FALSE}
# Incorrect: read_table will not correctly handle values that include spaces 
# such as "Flower Grove" in the school column.
choco_3 <- read_table("Chocolate Study 3.txt")
```

```{r eval=FALSE}
# Incorrect: This code does not tell R how to correctly identify missing values in the raw data.
choco_3 <- read_fwf(
  "Chocolate Study 3.txt",
  col_positions = fwf_widths(
    widths    = c(5, 9, 8, 4, 6, 13, 9, 9, 6, 6, 6, 6, 6, 6, 8, 7, 7, 7, 7, 9, 6),
    col_names = c(
      "uin", "last", "first", "age", "grade", "school", "parent", "marry", 
      "math1", "math2", "math3", "read1", "read2", "read3", "overall", "choco1", 
      "choco2", "choco3", "extra1", "extra2", "veggie"
    )
  ),
  skip = 1,
  na.rm = TRUE
)
```

```{r eval=FALSE}
# Incorrect: This code gives column start values instead of column widths
choco_3 <- read_fwf(
  "Chocolate Study 3.txt",
  col_positions = fwf_widths(
    widths    = c(5, 9, 8, 4, 6, 13, 9, 9, 6, 6, 6, 6, 6, 6, 8, 7, 7, 7, 7, 9, 6),
    col_names = c(
      "uin", "last", "first", "age", "grade", "school", "parent", "marry", 
      "math1", "math2", "math3", "read1", "read2", "read3", "overall", "choco1", 
      "choco2", "choco3", "extra1", "extra2", "veggie"
    )
  ),
  skip = 1,
  na = "."
)
```

```{r eval=FALSE}
# Correct
choco_3 <- read_fwf(
  "Chocolate Study 3.txt",
  col_positions = fwf_widths(
    widths    = c(5, 9, 8, 4, 6, 13, 9, 9, 6, 6, 6, 6, 6, 6, 8, 7, 7, 7, 7, 9, 6),
    col_names = c(
      "uin", "last", "first", "age", "grade", "school", "parent", "marry", 
      "math1", "math2", "math3", "read1", "read2", "read3", "overall", "choco1", 
      "choco2", "choco3", "extra1", "extra2", "veggie"
    )
  ),
  skip = 1,
  na = "."
)
```

```{r}
# For Dr. Cannell's computer.
choco_3 <- read_fwf(
  "/Users/bradcannell/Dropbox/Datasets/Chocolate/Chocolate Study 3.txt",
  col_positions = fwf_widths(
    widths    = c(5, 9, 8, 4, 6, 13, 9, 9, 6, 6, 6, 6, 6, 6, 8, 7, 7, 7, 7, 9, 6),
    col_names = c(
      "uin", "last", "first", "age", "grade", "school", "parent", "marry", 
      "math1", "math2", "math3", "read1", "read2", "read3", "overall", "choco1", 
      "choco2", "choco3", "extra1", "extra2", "veggie"
    )
  ),
  skip = 1,
  na = "."
)
```

# Task 7 

View the structure of `choco_3` using the `str()` function.

```{r}
str(choco_3)
```

## Question:

6. The `choco_3` data frame contains ____ rows and ____ columns.

* The `choco_3` data frame contains 10 rows and 21 columns.

# Task 8. 

Please view the National Health and Nutrition Examination Survey (NHANES) questionnaires, datasets, and related documentation website here: https://wwwn.cdc.gov/nchs/nhanes/ 

# Task 9. 

Import the 2015-2016 Alcohol Use data. Alcohol use was part of the NHANES questionnaire data.

```{r}
alq_i <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/ALQ_I.XPT")
```

## Question:

7. What R package contains the `read_xpt()` function we need to use to import the NHANES data in Task 9?

* The [haven](https://haven.tidyverse.org/) package contains the `read_xpt()` function we need to use to import the NHANES data in Task 9.

# Task 10

View the structure of `alq_i` using the `str()` function.

```{r}
str(alq_i)
```

## Questions:

9. The `alq_i` data frame contains ____ rows and ____ columns.

* The `alq_i` data frame contains 5,735 rows and 10 columns.

10. When you view the structure of `alq_i` using the `str()` function, what is the vector type shown for `alq101_f`? According to R, what kind of variable is `alq101_f`? 

* When you view the structure of `alq_i` using the `str()` function, the vector type shown is `Factor`.

