---
title: "Lab Warm-up: Cohort Studies"
format:
  html:
    embed-resources: true
---

# 🛑 Formatting (delete this section before posting to Canvas) 🛑

* Please name this file `lab_warm_up_module_name.qmd`.
* The Overview section, each task number, and each question number should be written as a level-one headers (e.g., `# Task 1`).
* All R code being run, should be put into R code chunks.
* All R code being discussed in the text should be placed inside backticks (e.g., `install.packages("dplyr")`).
* R package names being discussed in the text should be formatted as code (e.g., "...the `dplyr` package...). 
* When functions are discussed in the text, they should be formatted as code and empty parentheses added to the end of the function name (e.g., "The `tibble()` function can be used to create data frames…").
* Data set/frame names, variable names, and other key words should be written with **bold** text (e.g., "Please download **chocolate study 1.xls**".
* There should be 2 line breaks (returns) immediately above level-one headers. There should be 1 line break (return) immediately above all other headers.
* There should be one line break (return) immediately above each code chunk. There should be at least 1 line break immediately below each code chunk. If the the code chunk is immediately followed by a level-one header, then there will be two line breaks.

# Load packages

Remember, it's considered a best practice to load all the packages that your file will use up at the top of the file. If you have not yet installed a package, you will need to do so by running `install.packages("package name")` in the R console. For example, to install the `dplyr` package, you would run `install.packages("dplyr")`. However, you typically **do not** want to type the code to install the package here in your Quarto file because you only need to install the package once on a given computer. Not every time you run your R code.

```{r}
#| label: load-packages
library(dplyr, warn.conflicts = FALSE) # The "warn.conflicts" part is optional
library(ggplot2, warn.conflicts = FALSE)
```


# Load helper function

```{r}
# Convert from data frame of counts to data frame of cases.
# `count_col` is the name of the column containing the counts
# From: https://cran.r-project.org/web/packages/DescTools/vignettes/TablesInR.pdf
countsToCases <- function(x, count_col = "Freq") {
    # Get the row indices to pull from x
    idx <- rep.int(seq_len(nrow(x)), x[[countcol]])

    # Drop count column
    x[[countcol]] <- NULL

    # Get the rows from x
    x[idx, ]
}
```


# Grain auger use and unintentional injuries

## Simulate data to accompany the example.

```{r}
augers_and_injuries <- tribble(
  ~id, ~auger, ~injury,
  1,   1,      0,
  2,   1,      1,
  3,   1,      0,
  4,   1,      1,
  5,   0,      0,
  6,   0,      0,
  7,   0,      1,
  8,   0,      0,
  9,   0,      0,
  10,  0,      1,
  11,  0,      0, 
  12,  0,      0
)
```

## Create a person-time plot to accompany the example

```{r}
# Simulate data for 10 hypothetical people
# From Lash TL, VanderWeel TJ, Haneuse S, Rothman KJ. Modern Epidemiology. fourth. Wolters Kluwer; 2021.
people <- 1:12
months <- 0:12
augers_and_injuries_pt <- tidyr::expand_grid(person = people, month = months)
rows_per_person <- length(months)

# Count from the beginning of each month.
# So, person 2 is pre-follow-up at the beginning of time 1 (month -01), at risk 
# at the beginning of time 2 (month 00), at risk at the beginning of time 3 
# (month 01), diseased at the beginning of time 4 (month 02), etc.
p01 <- rep("At Risk", rows_per_person)
p02 <- c(rep("At Risk", 7), rep("Injured", rows_per_person - 7))
p03 <- rep("At Risk", rows_per_person)
p04 <- c(rep("At Risk", 2), rep("Injured", rows_per_person - 2))
p05 <- rep("At Risk", rows_per_person)
p06 <- rep("At Risk", rows_per_person)
p07 <- c(rep("At Risk", 5), rep("Injured", rows_per_person - 5))
p08 <- rep("At Risk", rows_per_person)
p09 <- rep("At Risk", rows_per_person)
p10 <- c(rep("At Risk", 10), rep("Injured", rows_per_person - 10))
p11 <- rep("At Risk", rows_per_person)
p12 <- rep("At Risk", rows_per_person)
status <- c(p01, p02, p03, p04, p05, p06, p07, p08, p09, p10, p11, p12)

# Add status to augers_and_injuries_pt
augers_and_injuries_pt$status <- status

# Create factor versions of person and status
augers_and_injuries_pt <- augers_and_injuries_pt |> 
  mutate(
    person_f = factor(person, labels = c(paste0(0, 1:9), "10", "11", "12")),
    status_f = factor(status, levels = c("At Risk", "Injured"))
  )

# Add exposure status
augers_and_injuries_pt <- augers_and_injuries_pt |> 
  mutate(
    # Persons 1 through 4 were exposed
    exposed   = if_else(person %in% 1:4, 1, 0),
    exposed_f = factor(exposed, 0:1, c("No", "Yes"))
  )

# Add coordinates for line segments in the plot
augers_and_injuries_pt_wide <- augers_and_injuries_pt |> 
  group_by(person_f, status_f) |> 
  mutate(
    x    = min(month),
    xend = max(month)
  ) |> 
  ungroup() |> 
  # Keep distinct values only
  distinct(person, status, .keep_all = TRUE) |> 
  # Make the line ends touch
  group_by(person) |> 
  mutate(
    xend_adjusted = lead(x),
    xend_adjusted = if_else(is.na(xend_adjusted), xend, xend_adjusted)
  ) |> 
  relocate(xend_adjusted, .after = xend) |> 
  ungroup()

# Print augers_and_injuries_pt_wide to the screen
# augers_and_injuries_pt_wide
```

```{r}
# To color the facet strip panels
stip_colors <- augers_and_injuries_pt_wide |> 
  # Keep one row for each person
  distinct(person, .keep_all = TRUE) |> 
  # Add color to use by exposed status
  mutate(color = if_else(exposed == 1, "#817190", "#D9D9D9")) |> 
  pull(color)
strips <- ggh4x::strip_themed(background_y = ggh4x::elem_list_rect(fill = stip_colors))

# Create the person-time figure
augers_and_injuries_plot <- ggplot(augers_and_injuries_pt_wide) +
  geom_segment(aes(x = x, xend = xend_adjusted, y = 1, yend = 1, color = status_f), size = 5) +
  # Show all months
  scale_x_continuous("Month Follow-up", breaks = 0:12) +
  ggh4x::facet_wrap2(vars(person_f), nrow = 12, strip.position = "left", strip = strips) +
  # Add a line marking the start of follow up
  geom_vline(xintercept = 0, color = "blue") +
  # Add vertical lines to mark the start of each month
  geom_vline(xintercept = 1:12, color = "#f9f9f9") +
  # Change fill colors
  scale_color_manual("Status", values = c("gray", "red")) +
  # Change y-axis label
  ylab("Person") +
  # Add a caption containing information about exposure status
  labs(caption = "Persons 1-4 were exposed to auger use.") +
  theme(
    # Remove gray background
    panel.background = element_blank(),
    # Remove the "p" from the x axis
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    # Orient Y axis title horizontally
    axis.title.y = element_text(angle = 0),
    # Move legend
    legend.position = "bottom",
    # Change the color and style of the "months" axis 
    axis.line.x = element_line(
      color = "blue", arrow = grid::arrow(length = unit(0.1, "inches"), type = "closed")
    ),
    # Rotate the facet label to the right
    strip.text.y.left = element_text(angle = 0),
    # Left justify the caption
    plot.caption = element_text(hjust = 0)
  )
```

```{r}
# View the person-time figure
augers_and_injuries_plot
```

```{r}
# Save the person-time figure as an image
ggsave("augers_and_injuries_plot.png", augers_and_injuries_plot, width = 6, height = 3.38)
```

## Create a contingency table

```{r}
# Create a contingency table called augers_and_injuries_ct
augers_and_injuries_ct <- matrix(
  c(a = 2, b = 2, c = 2, d = 6),
  ncol = 2,
  byrow = TRUE
)

# Add marginal totals the augers_and_injuries_ct
augers_and_injuries_ct <- addmargins(augers_and_injuries_ct)

# Assign row and column labels to augers_and_injuries_ct
dimnames(augers_and_injuries_ct) <- list(
  Auger = c("Yes", "No", "col_sum"), # Row names
  Injury  = c("Yes", "No", "row_sum")  # Then column names
)

# Print the value stored in augers_and_injuries_ct to the screen
augers_and_injuries_ct
```

## Incidence proportion

```{r}
# Calculate the incidence proportion by dividing each of the values in the "Yes"
# column of the contingency table by each of the values in the "row_sum" column
# of the contingency table. 
# This is equivalent to calculating:
# a / (a + b)
# c / (c + d)
# (a + c) / (a + b + c + d)
incidence_prop <- augers_and_injuries_ct[, "Yes"] / augers_and_injuries_ct[, "row_sum"]
# Print the value stored in incidence_prop to the screen
incidence_prop
```

```{r}
# Calculate incidence using R as a calculator
2 / 4
2 / 8
4 / 12
```


```{r}
# Add the incidence proportions we calculated above to the right side of our 
# contingency table
augers_and_injuries_ct <- cbind(augers_and_injuries_ct, incidence_prop)
# Print the value stored in augers_and_injuries_ct to the screen
augers_and_injuries_ct
```

## Incidence proportion ratio

```{r}
# Calculate the incidence proportion ratio and store it as a numeric vector 
# called ipr
ipr <- augers_and_injuries_ct["Yes", "incidence_prop"] / augers_and_injuries_ct["No", "incidence_prop"]
# Print the value stored in ipr to the screen
ipr
```

```{r}
# Calculate IPR using R as a calculator
0.5 / 0.25
```

Among the agricultural workers in or our study sample, those who used grain augers had 2 times the incidence of unintentional injury compared to those who did not use a grain auger, over the 1-year of follow-up.

## Incidence proportion difference

```{r}
# Calculate the incidence proportion difference and store it as a numeric vector 
# called ipd
ipd <- augers_and_injuries_ct["Yes", "incidence_prop"] - augers_and_injuries_ct["No", "incidence_prop"]
# Print the value stored in ipd to the screen
ipd
```

```{r}
# Calculate IPD using R as a calculator
0.5 - 0.25
```

Among the agricultural workers in or our study sample, there were an additional 0.25 incident cases of unintentional injury per person among workers who ever used a grain auger compared to those who never used a grain auger, over the 1-year of follow-up.

## Incidence odds

```{r}
# Add the incidence odds to augers_and_injuries_ct
incidence_odds <- augers_and_injuries_ct[, "incidence_prop"] / (1 - augers_and_injuries_ct[, "incidence_prop"])
augers_and_injuries_ct <- cbind(augers_and_injuries_ct, incidence_odds)
# Print the value stored in augers_and_injuries_ct to the screen
augers_and_injuries_ct
```

```{r}
# Calculate odds using R as a calculator
0.5 / (1 - 0.5)
0.25 / (1 - 0.25)
0.33 / (1 - 0.33)
```

```{r}
# Alternative odds calculation using R as a calculator
2 / 2
2 / 6
4 / 8
```

## Incidence odds ratio

```{r}
# Calculate the incidence odds ratio and store it as a numeric vector 
# called ior
ior <- augers_and_injuries_ct["Yes", "incidence_odds"] / augers_and_injuries_ct["No", "incidence_odds"]
# Print the value stored in ior to the screen
ior
```

```{r}
# Calculate OR using R as a calculator
1 / 0.3333333
```

Among the agricultural workers in or our study sample, those who used grain augers had 2 times the odds of unintentional injury compared to those who did not use a grain auger, over the 1-year of follow-up.

## Time at risk

```{r}
# Have R calculate the total person-months at risk
person_months <- augers_and_injuries_pt_wide |> 
  # Keep only the rows with a status of At Risk
  filter(status == "At Risk") |> 
  # Calculate the time between follow-up months for each person
  mutate(time_at_risk = xend_adjusted - x) |> 
  # Total all of the individual person-months at risk by exposure status
  group_by(exposed_f) |> 
  summarise(time_at_risk = sum(time_at_risk)) |> 
  # Make "Yes" appear first in the results
  arrange(desc(exposed_f)) %>% 
  # Add a total row
  # Make sure to use the magrittr pipe above
  add_row(exposed_f = "Total", time_at_risk = sum(.$time_at_risk))

# Print the value stored in augers_and_injuries_ct to the screen
person_months
```

```{r}
# Calculate time at risk using R as a calculator
(12 * 2) + 7 + 2
(12 * 6) + 5 + 10
```

```{r}
33 + 87
```

## Incidence rates

```{r}
# Add person months to the contingency table
augers_and_injuries_ct <- cbind(augers_and_injuries_ct, person_months = person_months$time_at_risk)
# Print the value stored in augers_and_injuries_ct to the screen
augers_and_injuries_ct
```

```{r}
# Add the incidence rates to augers_and_injuries_ct
incidence_rate <- augers_and_injuries_ct[, "Yes"] / augers_and_injuries_ct[, "person_months"]
augers_and_injuries_ct <- cbind(augers_and_injuries_ct, incidence_rate)
# Print the value stored in augers_and_injuries_ct to the screen
augers_and_injuries_ct
```

```{r}
# Calculate incidence rate using R as a calculator
2 / 33
2 / 87
4 / 120
```

## Incidence rate ratio

```{r}
# Calculate the incidence rate ratio and store it as a numeric vector 
# called irr
irr <- augers_and_injuries_ct["Yes", "incidence_rate"] / augers_and_injuries_ct["No", "incidence_rate"]
# Print the value stored in irr to the screen
irr
```

```{r}
# Calculate IRR (rounded) using R as a calculator
0.06 / 0.02
```

Among the agricultural workers in or our study sample, those who used grain augers had 2.6 times the rate of unintentional injury compared to those who did not use a grain auger, over the 1-year of follow-up.

## Incidence rate difference

```{r}
# Calculate the incidence rate difference and store it as a numeric vector 
# called ird
ird <- augers_and_injuries_ct["Yes", "incidence_rate"] - augers_and_injuries_ct["No", "incidence_rate"]
# Print the value stored in ird to the screen
ird
```

```{r}
# Calculate IRD (rounded) using R as a calculator
0.06 - 0.02
```

Among the agricultural workers in or our study sample, there were an additional 0.04 incident cases of unintentional injury per person-month among workers who ever used a grain auger compared to those who never used a grain auger, over the 1-year of follow-up.







