---
title: "Survival Analysis for Cohort Studies II Module"
date: "2020-10-30"
---

```{r}
library(dplyr)
library(haven)
library(ggplot2)
library(forcats)
library(tidyr)
library(purrr)
library(survival)
library(survminer)
library(broom)
library(meantables)
```

# Load data

```{r}
bewick <- read_dta("/Users/bradcannell/Dropbox/Datasets/Bewick data/Bewick 2004 Table 1.dta")
```

# Example from Szklo and Nieto page 58 and 59

```{r}
sn_chap_2 <- tibble(
  person_no = 1:10,
  months = c(1, 17, 20, 9, 24, 16, 2, 13, 10, 3),
  outcome = c(1, 1, 1, 1, 2, 2, 2, 1, 2, 1),
  event = c(1, 1, 1, 1, 0, 0, 0, 1, 0, 1),
  censor = c(0, 0, 0, 0, 1, 1, 1, 0, 1, 0),
  person_no_f = factor(person_no),
  outcome_f = factor(outcome, 1:2, c("Death", "Censored")),
  event_f = factor(event, 0:1, c("No", "Yes")),
  censor_f = factor(censor, 0:1, c("No", "Yes"))
) %>% 
  print()
```

Figure 2-2 - page 55

```{r}
ggplot(sn_chap_2, aes(fct_rev(person_no_f), months)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  geom_point(
    aes(person_no_f, months, color = outcome_f, shape = outcome_f), 
    size = 6
  ) +
  coord_flip() +
  labs(
    x = "Person Number",
    y = "Time (Months)"
  ) +
  theme_minimal() + 
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

Figure 2-3 - page 58

```{r}
ggplot(sn_chap_2, aes(fct_reorder(person_no_f, months, .desc = TRUE), months)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  geom_point(
    aes(person_no_f, months, color = outcome_f, shape = outcome_f), 
    size = 6
  ) +
  coord_flip() +
  labs(
    x = "Person Number",
    y = "Time (Months)"
  ) +
  theme_minimal() + 
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

Table 2-3. Calculation of Kaplan-Meier survival estimates for the example in Figure 2-3. Page 59.

```{r}
tab_2_3 <- sn_chap_2 %>% 
  group_by(months) %>% 
  summarise(
    n = n(),
    events = sum(event),
    censored = sum(censor),
    .groups = "drop"
  ) %>% 
  ungroup() %>% 
  complete(
    months = seq(1, max(sn_chap_2$months)),
    fill = list(n = 0, events = 0, censored = 0)
  ) %>% 
  mutate(
    # This is clunky. I think I can do better.
    n_at_risk_start = max(cumsum(n)),
    cum_events = cumsum(events),
    cum_censored = cumsum(censored),
    n_at_risk = n_at_risk_start - lag(cum_events) - lag(cum_censored),
    n_at_risk = if_else(is.na(n_at_risk), n_at_risk_start, n_at_risk)
  ) %>% 
  select(
    months, n_at_risk, n_events = events, n_censored = censored
  ) %>% 
  mutate(
    cond_prob_event = round(n_events / n_at_risk, 3),
    cond_prob_survival = 1 - cond_prob_event
  ) %>% 
  print()
```

```{r}
out <- vector(mode = "numeric", 24)
for (i in seq_along(tab_2_3$cond_prob_survival)) {
  if (i == 1) {
    out[i] <- tab_2_3$cond_prob_survival[i]
  } else {
    out[i] <- out[i - 1] * tab_2_3$cond_prob_survival[i]
  }
}

tab_2_3$cum_prob_survival <- out
tab_2_3
```

```{r}
# Only show rows with event to match book
tab_2_3 %>% 
  filter(!n_events == 0)
```

# For SO: 

Calculate cumulative probability (Kaplan-Meier survival function) in a `dplyr` pipeline. 

I'm trying to create Kaplan-Meier life tables using a `dplyr` pipeline. I'm having trouble calculating the cumulative probability of survival without using a for loop. Here is some example data. 

```{r}
df <- tibble(
  person_no = 1:10,
  months = c(1, 17, 20, 9, 24, 16, 2, 13, 10, 3),
  outcome = c(1, 1, 1, 1, 2, 2, 2, 1, 2, 1),
  event = c(1, 1, 1, 1, 0, 0, 0, 1, 0, 1),
  censor = c(0, 0, 0, 0, 1, 1, 1, 0, 1, 0),
  person_no_f = factor(person_no),
  outcome_f = factor(outcome, 1:2, c("Death", "Censored")),
  event_f = factor(event, 0:1, c("No", "Yes")),
  censor_f = factor(censor, 0:1, c("No", "Yes"))
) %>% 
  print()
```

```{r}
# Don't actually submit this. It's just for getting the code. 
df %>% 
  group_by(months) %>% 
  summarise(
    n = n(),
    events = sum(event),
    censored = sum(censor),
    .groups = "drop"
  ) %>% 
  ungroup() %>% 
  # complete(
  #   months = seq(1, max(sn_chap_2$months)),
  #   fill = list(n = 0, events = 0, censored = 0)
  # ) %>% 
  mutate(
    n_at_risk_start = max(cumsum(n)),
    cum_events = cumsum(events),
    cum_censored = cumsum(censored),
    n_at_risk = n_at_risk_start - lag(cum_events) - lag(cum_censored),
    n_at_risk = if_else(is.na(n_at_risk), n_at_risk_start, as.integer(n_at_risk))
  ) %>% 
  select(
    months, n_at_risk, n_events = events, n_censored = censored
  ) %>% 
  mutate(
    cond_prob_event = round(n_events / n_at_risk, 3),
    cond_prob_survival = 1 - cond_prob_event
  ) %>% 
  filter(!n_events == 0)
```

```{r}
df <- tibble(
  months = c(1, 3, 9, 13, 17, 20),
  n_at_risk = c(10, 8, 7, 5, 3, 2),
  cond_prob_event = c(0.100, 0.125, 0.143, 0.200, 0.333, 0.500),
  cond_prob_surv = c(0.900, 0.875, 0.857, 0.800, 0.667, 0.50)
)

df
```

In this case the cumulative probability of survival is calculated as the product of the previous (lagged) cumulative probability of survival and the current conditional probability of survival. I can get the answer I'm looking for using a for loop:


```{r}
out <- vector(mode = "numeric", 6)

for (i in seq_along(df$cond_prob_surv)) {
  if (i == 1) {
    out[i] <- df$cond_prob_surv[i]
  } else {
    out[i] <- out[i - 1] * df$cond_prob_surv[i]
  }
}

df$cum_prob_survival <- out
df
```

However, I'd really like to find a `dplyr` only solution. Any help is greatly appreciated!

```{r}
df %>% 
  mutate(
    cum_prob_survival = 0.9,
    cum_prob_survival = map2_dbl(
      cond_prob_surv, 
      lag(cum_prob_survival),
      ~ .x * .y
    )
  )
```

```{r}
df %>% 
  mutate(
    cum_prob_survival = 0.9,
    cum_prob_survival = accumulate(cond_prob_surv, `*`)
  )
```




# Data management

```{r}
bewick <- bewick %>% 
  mutate(
    patientid_f = factor(patientid),
    treatment_f = factor(treatment),
    event_f = factor(event, 0:1, c("No", "Yes")),
    censor_f = factor(censor, 0:1, c("No", "Yes"))
  )
```


# plot with shapes to indicate censoring or event

```{r}
ggplot(bewick, aes(patientid_f, time_days)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  geom_point(
    aes(patientid_f, time_days, color = censor_f, shape = censor_f), 
    size = 6
  ) +
  coord_flip() +
  labs(
    x = "Patient ID",
    y = "Time (Days)"
  ) +
  theme_minimal() + 
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

# Create a Kaplan-Meier life table

Based on Szklo Table 2-3

```{r}
bewick %>% 
  group_by(time_days) %>% 
  summarise(
    n       = n(),
    events  = sum(event),
    .groups = "drop"
  ) %>% 
  mutate(
    n_at_risk   = sort(cumsum(n) - events, decreasing = TRUE),
    cond_prob_e = events/n_at_risk,
    cond_prob_s = (n_at_risk - events) / n_at_risk,
    cum_prob_s  = cumprod(cond_prob_s) 
  ) %>% 
  # Improve appearance
  mutate(
    across(
      cond_prob_e:cum_prob_s,
      round, 3
    )
  ) %>% 
  select(time_days, n_at_risk, events, cond_prob_e:cum_prob_s)
```

```{r}
(f1 <- survfit(Surv(time_days, event) ~ 1, bewick))
```

```{r}
ggsurvplot(f1, data = bewick)
```

```{r}
broom::tidy(f1) %>% summary()
```

```{r}
(f2 <- survfit(Surv(time_days, event) ~ treatment, bewick))
```

```{r}
ggsurvplot(f2)
```

```{r}
survdiff(Surv(time_days, event) ~ treatment, bewick)
```

# Recreate the analyses I used in the lab-warmup slides. 

I originally took them from the Regression Methods in Biostatistics Textbook.

## Acute lymphoblastic leukemia (ALL) data

```{r}
leuk <- read_dta("/Users/bradcannell/Dropbox/Datasets/Regression Methods in Biostatistics Textbook/data/leuk.dta")
```

## Data management

Cens is really event. Look at placebo group, they are all 1's.

Time is in weeks.

```{r}
leuk <- leuk %>% 
  mutate(
    # Cens is really event. Look at placebo group, they are all 1's
    event     = cens,
    censored  = if_else(event == 1, 0, 1),
    group_f   = if_else(group == 1, "6-MP", "Placebo")
  ) %>% 
  select(-cens, - group)
```

## View data

Time is in weeks.

```{r}
leuk
```

## Naive comparison of mean survival times

```{r}
leuk %>% 
  group_by(group_f) %>% 
  mean_table(time)
```

## Life table - placebo

```{r}
bewick %>% 
  group_by(time_days) %>% 
  summarise(
    n       = n(),
    events  = sum(event),
    .groups = "drop"
  ) %>% 
  mutate(
    n_at_risk   = sort(cumsum(n) - events, decreasing = TRUE),
    cond_prob_e = events/n_at_risk,
    cond_prob_s = (n_at_risk - events) / n_at_risk,
    cum_prob_s  = cumprod(cond_prob_s) 
  ) %>% 
  # Improve appearance
  mutate(
    across(
      cond_prob_e:cum_prob_s,
      round, 3
    )
  ) %>% 
  select(time_days, n_at_risk, events, cond_prob_e:cum_prob_s)
```

This is really clunky. I think I can do better. Maybe write my own function.

This stuff isn't correct yet

```{r}
# leuk %>% 
#   filter(group_f == "Placebo") %>% 
#   group_by(time) %>% 
#   summarise(
#     n        = n(),
#     events   = sum(event),
#     censored = sum(censored),
#     .groups = "drop"
#   ) %>% 
#   tidyr::complete(
#     time = seq(1, max(leuk$time)), 
#     fill = list(n = 0, events = 0, censored = 0)
#   ) %>%
#   mutate(
#     # This is clunky. I think I can do better.
#     n_at_risk_start = max(cumsum(n)),
#     cum_events = cumsum(events),
#     n_at_risk = n_at_risk_start - lag(cum_events),
#     n_at_risk = if_else(is.na(n_at_risk), n_at_risk_start, n_at_risk)
#   ) %>% 
#   select(
#     week = time, n_followed = n_at_risk, n_relapsed = events, n_censored = censored
#   ) %>% 
#   mutate(
#     n_remission = n_followed - n_relapsed,
#     cond_prob_remission = n_remission / n_followed,
#     survival_function = lag(cond_prob_remission) * cond_prob_remission
#   )
```

```{r}
# leuk %>% 
#   filter(group_f == "6-MP") %>% 
#   group_by(time) %>% 
#   summarise(
#     n        = n(),
#     events   = sum(event),
#     censored = sum(censored),
#     .groups = "drop"
#   ) %>% 
#   tidyr::complete(
#     time = seq(1, max(leuk$time)), 
#     fill = list(n = 0, events = 0, censored = 0)
#   ) %>%
#   mutate(
#     # This is clunky. I think I can do better.
#     n_at_risk_start = max(cumsum(n)),
#     cum_events = cumsum(events),
#     n_at_risk = n_at_risk_start - lag(cum_events),
#     n_at_risk = if_else(is.na(n_at_risk), n_at_risk_start, n_at_risk)
#   ) %>% 
#   select(
#     week = time, n_followed = n_at_risk, n_relapsed = events, n_censored = censored
#   ) %>% 
#   mutate(
#     n_remission = n_followed - n_relapsed,
#     cond_prob_remission = n_remission / n_followed,
#     survival_function = lag(cond_prob_remission)
#   )
```

```{r}
leuk %>%
  filter(group_f == "6-MP") %>% 
  filter(time <= 10) %>% 
  arrange(time)
```







































