---
title: "Lab: Causal Inference"
format:
  html:
    embed-resources: true
---

# Load packages

Remember, it's considered a best practice to load all the packages that your file will use up at the top of the file. If you have not yet installed a package, you will need to do so by running `install.packages("package name")` in the R console. For example, to install the `dplyr` package, you would run `install.packages("dplyr")`. However, you typically **do not** want to type the code to install the package here in your Quarto file because you only need to install the package once on a given computer. Not every time you run your R code.

```{r}
library(dplyr, warn.conflicts = FALSE)
library(dagitty)
```


# Overview

The following lab will cover content that was reviewed in the readings and lectures, including causal models (sufficient component causes and directed acyclic graphs) and a short review of descriptive analysis and regression techniques.


# Task 1

Use R to construct a DAG that depicts a causal relationship between `smoking` and `MI` (myocardial infarctions)? `Smoking` is the cause and `MI` is the effect.

```{r}
dag_1 <- dagitty('dag {
  Smoking [pos="0,1"]
  MI [pos="1,1"]
  Smoking -> MI
}')
plot(dag_1)
```


# Task 2

Add `cholesterol` to the DAG you created in task 1. In this case, `smoking` is a cause of `cholesterol` and `cholesterol` is a cause of `MI`.

```{r}
dag_2 <- dagitty('dag {
  Smoking [pos="0,1"]
  Cholesterol [pos="1,1"]
  MI [pos="2,1"]
  Smoking -> Cholesterol -> MI
}')
plot(dag_2)
```


## Question 1

In the final DAG model, `smoking` is the [exposure, mediator, outcome].

* In the final DAG model, `smoking` is the exposure.


## Question 2

In the final DAG model, `cholesterol` is the [exposure, mediator, outcome].

* In the final DAG model, `cholesterol` is the mediator.


## Question 3

In the final DAG model, `MI` is the [exposure, mediator, outcome].

* In the final DAG model, `MI` is the outcome.



