---
title: "Lab Warm-up: Module Name"
output: html_notebook
---

# Load packages

Remember, it's considered a best practice to load all the packages that your file will use up at the top of the file. If you have not yet installed a package, you will need to do so by running `install.packages("package name")` in the R console. For example, to install the `dplyr` package, you would run `install.packages("dplyr")`. However, you typically **do not** want to type the code to install the package here in your Rmd file. That is because you only need to install the package once on a given computer. Not every time you run your R code.

```{r}
library(dplyr, warn.conflicts = FALSE) # The "warn.conflicts" part is optional
library(dagitty)
library(ggdag, warn.conflicts = FALSE)
```

# Overview

We are learning about causal inference in this module, and we are learning how to use directed acyclic graphs (DAGs) as a tool that can help us reason through causal inference. It turns out that there is also an R package that can help us create and interpret DAGs -- `dagitty`. In the code below, we will introduce `dagitty` and walk through some of the package's basic functionality together.

Here is a [link to a useful website](http://www.dagitty.net/) for learning more about `dagitty`.

Here is a [link to a useful website](https://cran.r-project.org/web/packages/ggdag/vignettes/intro-to-ggdag.ht) for learning more about `ggdag`.


# Create a DAG in the web browser

The easiest way to get started with using `dagitty` is arguably to draw our DAGs using `dagitty` in our web browser. You can do that by navigating to http://www.dagitty.net/dags.html.

To start with a blank slate, click **Model**, then **New Model**.

To add a node to the graph, click anywhere on the graph space and name your new node.

To connect nodes, first click the **cause** node and then click the **effect** node.

To save your DAG code, copy it from the **Model code** pane on the right side of the screen. Then, paste that code into an R code chunk. Just make sure you wrap the code you paste from the web browser in single quotes and place it inside the `daggity()` function.


# Create a DAG with R code

## Chain

```{r}
dag <- dagitty('dag {
  bb="0,0,1,1"
  x [pos="0.124,0.363"]
  y [pos="0.391,0.363"]
  z [pos="0.638,0.363"]
  x -> y
  y -> z
}')
```

```{r}
plot(dag)
```

```{r}
ggdag(dag)
```

## Collider 

```{r}

```

## Common cause

```{r}

```

## Common effect

```{r}

```

