---
title: "Regression Lecture"
date: "2020-12-08"
---

```{r}
library(dplyr)
library(haven)
library(broom)
```

# Load data

```{r}
lowbt <- read_dta("/Users/bradcannell/Dropbox/02 Teaching/Courses/Epi 3/modules/12 Cross-sectional II/lowbwt.dta")
```

# Linear regression

## Regress mother's weight on mother's age (continuous predictor)

```{r}
lm1 <- lm(lwt ~ age, data = lowbt)
tidy(lm1, conf.int = TRUE)
```

## Regress mother's race on mother's race (categorical predictor)

```{r}
lm2 <- lm(lwt ~ factor(race), data = lowbt)
tidy(lm2, conf.int = TRUE)
```

## Regress mother's weight on mother's age and mother's race (multiple predictors)

```{r}
lm3 <- lm(lwt ~ age + factor(race), data = lowbt)
tidy(lm3, conf.int = TRUE)
```


# Logistic regression

## Regress low birth weight on mother's age (continuous predictor)

```{r}
logistic1 <- glm(low ~ age, family = "binomial", data = lowbt)
tidy(logistic1, conf.int = TRUE)
```

```{r}
# For odds ratios
tidy(logistic1, conf.int = TRUE, exponentiate = TRUE)
```

## Regress low birth weight on mother's race (categorical predictor)

```{r}
logistic2 <- glm(low ~ factor(race), family = "binomial", data = lowbt)
tidy(logistic2, conf.int = TRUE)
```

```{r}
# For odds ratios
tidy(logistic2, conf.int = TRUE, exponentiate = TRUE)
```

## Regress low birth weight on mother's age and mother's race (multiple predictors)

```{r}
logistic3 <- glm(low ~ age + factor(race), family = "binomial", data = lowbt)
tidy(logistic3, conf.int = TRUE)
```

```{r}
# For odds ratios
tidy(logistic3, conf.int = TRUE, exponentiate = TRUE)
```
