---
title: "Lab Warm-up: Module Name"
format: html
---

# 🛑 Formatting (delete this section before posting to Canvas) 🛑

* Please name this file `lab_warm_up_module_name.qmd`.
* The Overview section, each task number, and each question number should be written as a level-one headers (e.g., `# Task 1`).
* All R code being run, should be put into R code chunks.
* All R code being discussed in the text should be placed inside backticks (e.g., `install.packages("dplyr")`).
* R package names being discussed in the text should be formatted as code (e.g., "...the `dplyr` package...). 
* When functions are discussed in the text, they should be formatted as code and empty parentheses added to the end of the function name (e.g., "The `tibble()` function can be used to create data frames…").
* Data set/frame names, variable names, and other key words should be written with **bold** text (e.g., "Please download **chocolate study 1.xls**".
* There should be 2 line breaks (returns) immediately above level-one headers. There should be 1 line break (return) immediately above all other headers.
* There should be one line break (return) immediately above each code chunk. There should be at least 1 line break immediately below each code chunk. If the the code chunk is immediately followed by a level-one header, then there will be two line breaks.

# Load packages

Remember, it's considered a best practice to load all the packages that your file will use up at the top of the file. If you have not yet installed a package, you will need to do so by running `install.packages("package name")` in the R console. For example, to install the `dplyr` package, you would run `install.packages("dplyr")`. However, you typically **do not** want to type the code to install the package here in your Quarto file because you only need to install the package once on a given computer. Not every time you run your R code.

```{r}
#| label: load-packages
library(dplyr, warn.conflicts = FALSE) # The "warn.conflicts" part is optional
```

# Memoral study

```{r}
# Create a contingency table
memoral <- matrix(
  c(a = 6, b = 7, c = 6, d = 6),
  ncol = 2,
  byrow = TRUE
)

# Assign row and column labels
dimnames(memoral) <- list(
  Exposure = c("Obese", "Non-obese"), # Row names
  Outcome  = c("Cases", "Controls")  # Then column names
)

# Add marginal totals to the contingency table
memoral <- addmargins(memoral)

# Assign row and column labels to the contingency table
dimnames(memoral) <- list(
  Memoral = c("Yes", "No", "col_sum"), # Row names
  `AD Symptoms`  = c("Yes", "No", "row_sum")  # Then column names
)

# Print the contingency table to the screen
memoral
```

```{r}
# Add the incidence proportions to the contingency table
incidence_prop <- memoral[, "Yes"] / memoral[, "row_sum"]
memoral <- cbind(memoral, incidence_prop)
# Print the contingency table to the screen
memoral
```

```{r}
# Calculate the incidence proportion ratio and store it as a numeric vector 
# called ipr
ipr <- memoral["Yes", "incidence_prop"] / memoral["No", "incidence_prop"]
# Print the ipr to the screen
ipr
```

Among the members of this study population, those who were assigned to the Memoral group had 0.92 times the incidence of Alzheimer's symptoms compared to those who were assigned to the control group over the study's follow-up period.

```{r}
# % decrease = (1 - RR) x 100
# Source: https://sphweb.bumc.bu.edu/otlt/mph-modules/ep/ep713_association/ep713_association3.html
(1 - ipr) * 100
```

Among the members of this study population, those who were assigned to the Memoral group had 8% reduction in incident Alzheimer's symptoms compared to those who were assigned to the control group over the study's follow-up period.

```{r}
# Calculate the incidence proportion difference and store it as a numeric vector 
# called ipd
ipd <- memoral["Yes", "incidence_prop"] - memoral["No", "incidence_prop"]
# Print the ipd to the screen
ipd
```

Among the members of this study population, those who were assigned to the Memoral group had 0.04 fewer cases of incident Alzheimer's symptoms per person compared to those who were assigned to the control group over the study's follow-up period.

```{r}
# Multiply the ipd by 100 to make the units 100 people
ipd * 100
```

Among the members of this study population, those who were assigned to the Memoral group had 4 fewer cases of incident Alzheimer's symptoms per 100 people compared to those who were assigned to the control group over the study's follow-up period.



