---
title: "Effect Modification"
---

```{r}
library(dplyr, warn.conflicts = FALSE)
library(freqtables)
```

Here's what I want to do:

* Input a contingency table

* Display the cases

* Convert from raw data to data frame

* Display contingency table as a data frame

* Display the contingency table as a 2x2 table (matrix)

Helpful websites:

* http://www.cookbook-r.com/Manipulating_data/Converting_between_data_frames_and_contingency_tables/

# Input a contingency table

Given a contingency table (e.g., from a paper or textbook), what's the best way to get that data into R and then start to calculate things like the OR, etc.?



# Display the cases

By cases, I mean typical data where each row is a person and the columns are exposure, outcome, and other relevant information.

With cases, it seems like either: 

1. The data would already be in a data set/data frame and nothing special would 

```{r}
data <- tibble(
  sex = rep(c("Female", "Male"), c(165, 135)),
  exposed = rep("Yes")
) %>% 
  print()
```


```{r}
hernan_4_1 <- tribble(
  ~Name, ~V, ~Y0, ~Y1,
  "Rheia", 1, 0, 1, 
  "Demeter", 1,0, 0,
  "Hestia", 1, 0, 0,
  "Hera", 1, 0, 0,
  "Artemis", 1, 1, 1,
  "Leto", 1, 0, 1,
  "Athena", 1, 1, 1,
  "Aphrodite", 1, 0, 1,
  "Persephone", 1, 1, 1,
  "Hebe", 1, 1, 0,
  "Kronos", 0, 1, 0,
  "Hades", 0, 0, 0,
  "Poseidon", 0, 1, 0,
  "Zeus", 0, 0, 1,
  "Apollo", 0, 1, 0,
  "Ares", 0, 1, 1,
  "Hephaestus", 0, 0, 1,
  "Cyclope", 0, 0, 1,
  "Hermes", 0, 1, 0,
  "Dionysus", 0, 1, 0
)
```

Half of the members of the population would have died ifeverybody had received a heart transplant,Pr[=1=1]=1020 = 05,and half of the members of the population would have died if nobody had re-ceived a heart transplant,Pr[=0=1]=1020 = 05. The causal risk ratioPr[=1=1]Pr[=0=1]was0505=1and the causal risk differencePr[=1=1]−Pr[=0=1]was05−05=0.

```{r}
hernan_4_1 %>% 
  summarise(
    Pr_Ya1_1 = mean(Y1),
    Pr_Ya0_1 = mean(Y0),
    RR = Pr_Ya1_1 / Pr_Ya0_1,
    RD = Pr_Ya1_1 - Pr_Ya0_1
  )
```

No causal effect of a (heart transplant) overall. 

We now consider two new causal questions: What is the average causaleffect ofonin women?  And in men?  To answer these questions wewill use Table 4.1, which contains the same information as Table 1.1 plus anadditional column with an indicatorfor sex:=1for females (referredto as women in this book) and=0for males (referred to as men).  Forconvenience, we have rearranged the table so that women occupy thefirst10rows, and men the last10rows.

```{r}
hernan_4_1 %>% 
  group_by(V) %>% 
  summarise(
    Pr_Ya1_1 = mean(Y1),
    Pr_Ya0_1 = mean(Y0),
    RR = Pr_Ya1_1 / Pr_Ya0_1,
    RD = Pr_Ya1_1 - Pr_Ya0_1,
    .groups = "drop"
  )
```
Now, heart transplant is protective in men (V=0) and harmful in women (V=1).



























